Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-10-29 manuscript/manuscript.Rmd
Adding DMS story
Adding phosphorus supplemental figure and discussion

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE)
library(tidyverse)
library(RaMS)
library(vegan)
library(cowplot)
library(viridis)
library(scales)
library(ggnewscale)
library(egg)
library(broom)
library(lme4)
library(openxlsx)
library(readxl)
library(ggh4x)
library(ggtext)
library(gridExtra)
library(openxlsx)

@@

@@

The HEE data was characterized by high inter-replicate variability relative to the samples from the MESO-SCOPE cruise, with 25 meter samples in particular highly variable in multivariate space (Figure 6). Despite this, we still saw the importance of sea level anomaly as a significant explanatory factor in the dataset (PERMANOVA R$^2_{SLA} = 0.090$, p-value = 0.017) as well as the larger depth differences between the surface (25 meters) and DCM (110 - 120 meters) (PERMANOVA R$^2_{depth} = 0.251$, p-value << 0.001). We saw no difference between the 6AM and 6PM samples (PERMANOVA R$^2_{time} = 0.031$, p-value = 0.367) despite a few known compounds (trehalose and sucrose) showing large differences, likely due to a majority of the mass features demonstrating no diel effect (Supplemental Figure 4).

@@

When analyzed separately as distinct depths, the 25 meter samples had higher variances explained by eddy and a lower p-value (PERMANOVA, R$^2$=0.26, p-value=0.013), while the DCM samples were no longer likely to be distinct between the two eddies (PERMANOVA p-value = 0.171) (Figure 6, Supplemental Figure 5).

@@

Isethionate is a known osmolyte thus far found exclusively in eukaryotes [@Durham2022] and strongly associated with a few diatoms in particular [@Heal2021]. This compound was enriched in the cyclone of the DCM in both the MESO-SCOPE and HEE cruises along with its precursor taurine, validating earlier results from @Barone2022 that showed strong enrichment of eukaryotic phytoplankton at the DCM and *Pseudo-nitzschia* in particular.

@@

*Supplemental Figure 3: Stacked barplots of known metabolite concentration as shown in main text Figure 3 but with individual triplicates shown. The same data is shown in A) as B) but B) is rendered in relative contribution space instead of absolute concentration. TMAO = trimethylamine N-oxide, HO-Ile = hydroxyisoleucine, GBT = glycine betaine, DCM = deep chlorophyll maximum.*

## Supp fig 4: Boxplots of selected compounds by depth, time, and SLA

@@

*Supplemental Figure 4: Boxplots showing the concentrations of six selected compounds between cruises and depths as well as time and SLA classes. The abscissa is separated by the time category into which sampling was binned, referring to 6-hour increments starting from 3AM (e.g. Midday contains casts collected between 9AM and 3PM). Colors denote the sea level anomaly bins with a ±5cm cutoff. For the eddy transect, six stations are included in the anticyclone, three in the cyclone, and two were neither. There were two morning stations, 4 midday stations, 2 evening stations, and 3 night stations. Diel effects are clearly visible for sucrose and trehalose while other compounds show very little diel differences. Trehalose was not quantified during the MESO-SCOPE transect for lack of an authentic standard to quantify the response factor.*

## Supp fig 5: HEE NMDS and PERMANOVA broken down by depth

@@

*Supplemental Figure 5: Plots of the metabolome during the Hawaiian Eddy Experiment, broken down by depth. The top row of plots are non-metric multidimensional scaling (NMDS) plots, in which points correspond to individual samples and have been colored by their corrected sea level height anomaly. SLA trends are visible in the 25 meter samples, with dark blue circles consistently discriminating from the dark red circles. The bottom row of plots show the direction and magnitude of this effect by plotting the grand mean of the normalized metabolite peak areas in the stations taken between the adjacent eddies of opposite polarity.*

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE)
library(tidyverse)
library(RaMS)
library(vegan)
library(cowplot)
library(viridis)
library(scales)
library(ggnewscale)
library(egg)
library(broom)
library(lme4)
library(openxlsx)
library(readxl)
library(ggh4x)
library(ggtext)
library(gridExtra)
library(openxlsx)
library(arrow)

@@

We also measured several phosphorus-containing compounds and their non-phosphorus equivalents (Supplemental Figure 4A). Phosphocholine is the polar headgroup for phosphatidylcholine, a class of intact polar diacylglycerols that has been previously described at this study site as showing phosphorus stress relief among eukaryotic organisms in response to the upwelling of phosphate-rich deep sea water [@Bent2024]. We measured this metabolite as well as choline (the non-phosphate containing equivalent) and glycerophosphocholine (GPC, which contains an additional glycerol group) with the expectation that phosphorus stress would result in higher choline:phosphocholine and choline:GPC ratios, as shown in Seelen et al. (in review). However, we saw no differences in this ratio across the eddy transect after accounting for depth effects (Supplemental Figure 4C). Additionally, choline:GPC ratios were highest at 175m where PO$_4^{3-}$ concentrations are highest, indicating that this metric may not accurately measure P-stress. Other phosphorus-containing lipid headgroups glycerophosphoglycerol and glycerophosphoethanolamine also showed no trend across the eddy at a given depth and instead showed trends functionally identical to the average metabolite in Figure 2 (Supplemental Figure 4B).

@@

The HEE data was characterized by high inter-replicate variability relative to the samples from the MESO-SCOPE cruise, with 25 meter samples in particular highly variable in multivariate space (Figure 6). Despite this, we still saw the importance of sea level anomaly as a significant explanatory factor in the dataset (PERMANOVA R$^2_{SLA} = 0.090$, p-value = 0.017) as well as the larger depth differences between the surface (25 meters) and DCM (110 - 120 meters) (PERMANOVA R$^2_{depth} = 0.251$, p-value << 0.001). We saw no difference between the 6AM and 6PM samples (PERMANOVA R$^2_{time} = 0.031$, p-value = 0.367) despite a few known compounds (trehalose and sucrose) showing large differences, likely due to a majority of the mass features demonstrating no diel effect (Supplemental Figure 5).

@@

When analyzed separately as distinct depths, the 25 meter samples had higher variances explained by eddy and a lower p-value (PERMANOVA, R$^2$=0.26, p-value=0.013), while the DCM samples were no longer likely to be distinct between the two eddies (PERMANOVA p-value = 0.171) (Figure 6, Supplemental Figure 6).

@@

Isethionate is a known osmolyte thus far found exclusively in eukaryotes [@Durham2022] and strongly associated with a few diatoms in particular [@Heal2021]. This compound was enriched in the cyclone of the DCM in both the MESO-SCOPE and HEE cruises along with its precursor taurine, validating earlier results from @Barone2022 that showed strong enrichment of eukaryotic phytoplankton at the DCM and *Pseudo-nitzschia* in particular. Additionally, the abundance other sulfur-containing compounds (e.g. dimethylsulfonioacetate and dimethylsulfoniopropionate) in the cyclone may indicate links between the formation of eukaryote-rich cyclonic features and the production of dimethylsulfide, a climate-active volatile gas [@Moran2019].

@@

*Supplemental Figure 4: Stacked barplots of known metabolite concentration as shown in main text Figure 3 but with individual triplicates shown. The same data is shown in A) as B) but B) is rendered in relative contribution space instead of absolute concentration. TMAO = trimethylamine N-oxide, HO-Ile = hydroxyisoleucine, GBT = glycine betaine, DCM = deep chlorophyll maximum.*

## Supp fig 4: Choline, phosphocholine, and glycerophosphocholine ratios

```{r}
gp_df <- data.frame(
  compound_name=c("Ethanolamine", "Glycerol", "Choline"),
  mz=c(62.060589, 93.05462, 104.107539)
) %>%
  expand_grid(data.frame(
    addition=c("", "Phospho", "Glycerophospho"),
    mz_diff=c(0, 79.96633, 74.03678+79.96633)
  )) %>%
  mutate(base_compound=compound_name) %>%
  mutate(compound_name=ifelse(addition=="", compound_name, 
                              paste0(addition, tolower(compound_name))),
         mz=mz+mz_diff) %>%
  arrange(mz) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  mutate(addition=fct_inorder(addition))
gp_w_eics <- gp_df %>%
  mutate(eic=map(mz, function(mz_i){
    open_dataset("~/My Drive/AllMeso/tmzMLs/MT_all_file_arr_pqds") %>%
      filter(mz%between%pmppm(mz_i, 10)) %>%
      collect()
  }))
smp_eics <- gp_w_eics %>% 
  select(-mz) %>% 
  unnest(eic) %>% 
  filter(str_detect(filename, "Smp")) %>%
  left_join(filled_file_metadata)
all_cmpds <- tribble(
  ~compound_name, ~rt_start, ~rt_end,
  "Choline", 10.6, 11.2,
  "Phosphocholine", 11.8, 13.2,
  "Glycerophosphocholine", 10.8, 11.5,
  "Glycerophosphoethanolamine", 11, 12.5,
  "Glycerophosphoglycerol", 10.2, 11.2
) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  left_join(smp_eics, by=join_by(compound_name, between(y$rt, x$rt_start, x$rt_end))) %>%
  left_join(filled_file_metadata) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), labels=c("15m", "DCM", "175m")))
```

```{r}
gpc_chrom_gp <- all_cmpds %>%
  mutate(compound_name=str_replace(compound_name, "Glycerophospho", "Glycerophospho-\n")) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  mutate(int=int/1e6) %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
  facet_wrap(~compound_name, scales="free", nrow=1) +
  scale_color_manual(values=viridis::viridis(9, direction = -1)[c(2,5,8)], 
                     breaks = c("15m", "DCM", "175m"),
                     aesthetics = c("color", "fill"),
                     name="Sample\ndepth") +
  theme_bw() +
  labs(x="Retention time (minutes)", y="Intensity (millions)") +
  guides(color=guide_legend(override.aes = list(lwd=2)))
gpc_area_gp <- all_cmpds %>%
  summarise(area=trapz(rt, int)/1e6, .by = c(compound_name, filename, depth, station, sla_corr)) %>%
  mutate(compound_name=str_replace(compound_name, "Glycerophospho", "Glycerophospho-\n")) %>%
  # mutate(station=factor(station, levels=as.character(4:14))) %>%
  mutate(station=as.numeric(station)) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=station, y=area, fill=sla_corr), color="black", pch=21) +
  facet_grid2(depth~compound_name, scales="free_y", independent = "y") +
  labs(x="Station number", y="Peak area (millions)", fill="Corr. SLA\n(cm)") +
  scale_color_gradient2(low = scales::muted("blue"), 
                        high = scales::muted("red"),
                        mid="grey90", aesthetics = c("color", "fill")) +
  theme_bw()

gpc_ratio_reg_gp <- all_cmpds %>%
  summarise(area=trapz(rt, int)/1e6, .by = c(compound_name, filename, depth, station, sla_corr)) %>%
  pivot_wider(names_from = compound_name, values_from=area) %>%
  pivot_longer(Phosphocholine:last_col()) %>%
  filter(name%in%c("Phosphocholine", "Glycerophosphocholine")) %>%
  mutate(name=paste0("Choline:", name)) %>%
  mutate(name=fct_inorder(name)) %>%
  ggplot(aes(x=sla_corr, y=Choline/value)) +
  geom_smooth(method="lm", formula=y~x, color="black") +
  geom_point(aes(fill=sla_corr), color="black", pch=21) +
  facet_nested(.~name+depth, scales="free_y", independent = "y") +
  labs(x="Corrected SLA (cm)", y="Metabolite ratio", fill="Corr. SLA\n(cm)") +
  scale_fill_gradient2(low = scales::muted("blue"), 
                        high = scales::muted("red"),
                        mid="grey90", guide = "none") +
  theme_bw()

comb_gpc_patch <- gpc_chrom_gp / gpc_area_gp / gpc_ratio_reg_gp + plot_annotation(tag_levels = 'A') + plot_layout(heights = c(0.3, 0.4, 0.3))
ggsave("supplement/comb_gpc_patch.png", comb_gpc_patch, device = "png", width = 9, height = 6.5, units = "in", dpi = 300, bg="white")
```

## Supp fig 5: Boxplots of selected compounds by depth, time, and SLA

@@

*Supplemental Figure 5: Boxplots showing the concentrations of six selected compounds between cruises and depths as well as time and SLA classes. The abscissa is separated by the time category into which sampling was binned, referring to 6-hour increments starting from 3AM (e.g. Midday contains casts collected between 9AM and 3PM). Colors denote the sea level anomaly bins with a ±5cm cutoff. For the eddy transect, six stations are included in the anticyclone, three in the cyclone, and two were neither. There were two morning stations, 4 midday stations, 2 evening stations, and 3 night stations. Diel effects are clearly visible for sucrose and trehalose while other compounds show very little diel differences. Trehalose was not quantified during the MESO-SCOPE transect for lack of an authentic standard to quantify the response factor.*

## Supp fig 6: HEE NMDS and PERMANOVA broken down by depth

@@

*Supplemental Figure 6: Plots of the metabolome during the Hawaiian Eddy Experiment, broken down by depth. The top row of plots are non-metric multidimensional scaling (NMDS) plots, in which points correspond to individual samples and have been colored by their corrected sea level height anomaly. SLA trends are visible in the 25 meter samples, with dark blue circles consistently discriminating from the dark red circles. The bottom row of plots show the direction and magnitude of this effect by plotting the grand mean of the normalized metabolite peak areas in the stations taken between the adjacent eddies of opposite polarity.*
